Joint estimation of ancestral states and phylogenetic model parameters is inconsistent

### 2017-11-13 joint_inf.tex
Re-outline.
update analytic plots (still no hatching)
hatch waves
update plot---will redo for y-axis
pass through until lower bound---need to reorganize

\title{On joint estimation of ancestral states and phylogenetic model parameters}
\author{Shaw, Matsen, Minin}

@@

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model and imagine data being generated on the four taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the large-data limit.

@@

We consider the standard phylogenetic likelihood on unrooted phylogenetic trees \cite{Felsenstein2004}.
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
Assume we observe $\nCols$ samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.

@@

An alternative approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
One can compute the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates),
\end{equation}
then estimate the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment),
\end{equation}
using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and \eqref{eq:profile_likelihood_topology_bl} recovers the joint ML values of the unknown parameters.
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.

@@

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
Indeed, we start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.

@@

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$.
For the other term, we define
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
as a mapping of ancestral states and tip states to ancestral state splits.
It is defined by if the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is not contained in $\siteSplitSet$ and thus we must take a complement to become a site split, we let $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ be the complement of $\ancestralStateColumn$; otherwise it is simply $\ancestralStateColumn$.
Note $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective for each $\alignmentColumn_i$.
We define
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
By considering the case requiring complementing and that not requiring complementing separately,
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
By the surjectivity of $\ancestralToSplit(\alignmentColumn_i, \cdot)$, we also have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

@@

\paragraph{Example}
\begin{figure}
    \centering
    \includegraphics[width=.45\textwidth]{unrooted_four_taxa}
    \caption{Four taxon tree}
\label{fig:four-taxa-tree}
\end{figure}

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{2\}$ and internal node $\{2\}$ connected to leaves $\{3\}$ and $\{4\}$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_6=\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2.
$$
Here, the set of most likely ancestral states is $\ancestralSplitPartition_6(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_6\in\ancestralSplitPartition_6(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_6 \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3.
$$

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Let
$$
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t))
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}
For data generated from $(\tau^*, t^*)$, we have an inconsistency if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\tau', \hat{t}_1; \fullAncestralSplitPartitions(\tau', \hat{t}_1)) > \ell_{\tau^*,t^*}(\tau^*, \hat{t}_2; \fullAncestralSplitPartitions(\tau^*, \hat{t}_2))
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\fullAncestralSplitPartitions(\tau', \hat{t}_1)\}$ and $\{\hat{t}_2,\fullAncestralSplitPartitions(\tau^*, \hat{t}_2)\}$ estimated by maximizing $\tilde{\ell}_{\tau^*,t^*}$ with respect to the branch lengths and the internal states for $\tau = \tau'$ and $\tau^*$, respectively.
By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
In the next section, we show that such examples exist.

@@

We will begin by writing out the middle case: for the ancestral state split $\{1\}$.

@@

    &\le      p_{\emptyset}  \cdot\log(2) \\

>>>>

\title{Joint estimation of ancestral states and phylogenetic model parameters is inconsistent}
\author{Shaw, Matsen, Minin}

@@

%EM clarify that we aren't doing sequence-level simulation somewhere.
In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model and imagine data being generated on the four taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the large-data limit.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known.

%EM Potential re-outline--
%
%EM 1. Describe setup with Fels and Farris trees, and how we are talking about the case of infinite data.
%EM 2. Describe fidelities
%EM 3. State theorems
%EM 4. Plots
%EM 5. Discussion, already below. This motivates a more thorough investigation under simulation.

@@

We consider the standard phylogenetic likelihood on unrooted trees \cite{Felsenstein2004}.
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
We observe $\nCols$ samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.

@@

An alternative approach does away with the marginalization and directly estimates the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
One can compute the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates),
\end{equation}
then estimate the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment),
\end{equation}
using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and \eqref{eq:profile_likelihood_topology_bl} recovers the joint ML values of the unknown parameters.
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.

@@

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
Indeed, we start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.

@@

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$.
For the other term, we define
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
as a mapping of ancestral states and tip states to ancestral state splits.
It is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
We define
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, so that
%By considering the case requiring complementing and that not requiring complementing separately,
%das: did I write the above? I don't remember the significance of it...
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

@@

\paragraph{Example}
\begin{figure}
    \centering
    \includegraphics[width=.45\textwidth]{unrooted_farris_tree}
    \caption{Four taxon ``Farris zone'' tree.}
\label{fig:four-taxa-tree}
\end{figure}

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}---this is commonly known as the ``Farris zone'' topology.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{3\}$ and internal node $\{2\}$ connected to leaves $\{2\}$ and $\{4\}$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_5=\{1,2\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
$$
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
$$

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Let
$$
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the limit of infinite data,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}
For data generated from $(\tau^*, t^*)$, we have an inconsistency in estimating topology if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\tau', \hat{t}_1; \fullAncestralSplitPartitions(\tau', \hat{t}_1)) > \ell_{\tau^*,t^*}(\tau^*, \hat{t}_2; \fullAncestralSplitPartitions(\tau^*, \hat{t}_2))
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\fullAncestralSplitPartitions(\tau', \hat{t}_1)\}$ and $\{\hat{t}_2,\fullAncestralSplitPartitions(\tau^*, \hat{t}_2)\}$ estimated by maximizing $\tilde{\ell}_{\tau^*,t^*}$ with respect to the branch lengths and the internal states for $\tau = \tau'$ and $\tau^*$, respectively.
By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
In the next section, we show that such examples exist.

@@

We begin by writing out the middle case: for the ancestral state split $\{1\}$.

@@

\begin{equation*}
\begin{split}
